---
title: "MinneMUDAC DS - Novice Questions"
subtitle: "Noah Lee"
output:
  pdf_document: default
  html_document: default
editor_options: 
  markdown: 
    wrap: 72
---

```{r}
#| include: false
library(tidyverse)
library(tidymodels)
library(naniar)    
library(dplyr)
library(ggformula)
library(ggplot2) 
library(GGally)

library(lubridate)
library(ranger)   
library(workflows) 
library(recipes) 
```

Big Brother Big Sisters Twin Cities the largest and oldest youth award-winning mentoring organization in the greater Twin Cities.  Each year, we match up youth (Littles age 8-13) and their families with caring adults (Bigs) who volunteer as mentors.  Through a variety of community-based, school-based, and workplace-based mentoring programs, and together with our community, we want every youth to have a mentor, be affirmed in who they are, and explore who they want to be.

# Question: What are things present in 'successful matches'? - matches ongoing, lasting long or closed 'successfully'?

Preprocessing: (THIS IS STUFF FROM MY PREVIOUS NOTEBOOKS - IGNORE OR LOOK BACK AT IT LATER - MAINLY DELETING IRRELEVANT COLUMNS AND DATA TRANSFORMATION)
```{r}
df <- read.csv('../Data/Novice.csv')
extract_binary_indicators <- function(df) {
  # Initialize new columns with FALSE (0)
  interest_categories <- c("has_interests", "personality_compatibility", "has_proximity", 
                          "has_commitment", "has_experience", "has_preference",
                          "has_challenges", "has_goals")
  
  for (category in interest_categories) {
    df[[category]] <- FALSE  # Initialize with FALSE for all rows
  }
  
  # Define keywords for each category
  keywords <- list(
    has_interests = c("outdoors", "hiking", "biking", "fishing", "camping", "parks", "nature", "gardening", "swimming", "sledding", "horseback riding", "skateboarding", "snowboarding", "ice skating", "picnics", "planting", "rock climbing", "feeding ducks", "flying kites", "sports", "basketball", "football", "soccer", "baseball", "hockey", "bowling", "tennis", "running", "yoga", "Zumba", "gymnastics", "arts", "crafts", "drawing", "painting", "pottery", "sewing", "knitting", "photography", "model cars", "model planes", "creative activities", "acting", "singing", "dancing", "playing musical instruments", "writing","indoor activities", "reading", "cooking", "baking", "board games", "video games", "puzzles", "Lego", "animals", "dogs", "cats", "horses", "pets", "animal care", "learning", "science", "math", "history", "social studies", "STEM projects", "other interests", "fashion", "hair", "nails", "volunteering", "museums", "libraries", "zoos", "movies", "plays"),
    personality_compatibility = c("outgoing", "talkative", "bubbly", "energetic", "enthusiastic", "charismatic", 
                                "shy", "reserved", "quiet", "introverted", "soft-spoken", "calm", 
                                "adventurous", "curious", "exploratory", "open to new things", 
                                "friendly", "kind", "sweet", "thoughtful", "empathetic", 
                                "funny", "goofy", "humorous", "light-hearted", 
                                "mature", "respectful", "responsible", "thoughtful", 
                                "active", "sporty", "energetic", "athletic", 
                                "creative", "imaginative", "artistic", "crafty", 
                                "patient", "calm", "steady", "nurturing"),
    has_proximity = c("miles", "minutes", "close", "far", "convenient", "driving", "traffic", "commute"),
    has_commitment = c("long-term", "committed", "consistent", "reliable", "short-term", "temporary", "limited time", "2-4 times a month", "weekly"),
    has_experience = c("child experience", "nanny", "teacher", "coach", "mentor", "social work", "counseling", "teaching", "overcoming challenges", "mental health"),
    has_preference= c("age", "younger", "older", "in 20s", "gender", "male", "female", "religion", "Christian", "Catholic", "cultural background", "African American", "Hispanic", "non-smoker", "no guns"),
    has_challenges = c("behavioral challenges", "ADHD", "unmedicated", "redirection", "mental health", "depression", "anxiety", "PTSD", "family dynamics", "divorce", "strained relationships", "bullying", "picked on", "self-esteem", "academic challenges", "tutoring", "homework help"),
    has_goals = c("self-esteem", "confidence", "self-image", "leadership", "decision-making", "independence", "academic success", "math", "science", "reading", "social skills", "communication", "making friends", "exploration", "trying new things", "learning new skills")
  )
  
  # Check if Rationale.for.Match column exists in the dataframe
  if (!"Rationale.for.Match" %in% names(df)) {
    warning("Column 'Rationale.for.Match' not found in dataframe. No keywords will be extracted.")
    # Return dataframe with all FALSE values
    return(df)
  }
  
  # Process each row
  for (i in 1:nrow(df)) {
    rationale <- df$Rationale.for.Match[i]
    
    # Skip if rationale is NA or empty
    if (is.na(rationale) || rationale == "") {
      next
    }
    
    # Check for keywords in each category
    for (category in names(keywords)) {
      category_keywords <- keywords[[category]]
      for (keyword in category_keywords) {
        if (grepl(keyword, rationale, ignore.case = TRUE)) {
          df[[category]][i] <- TRUE
          break  # Once we find a match, no need to check other keywords in this category
        }
      }
    }
  }
  
  # Convert logical columns to factors (0/1)
  for (category in interest_categories) {
    df[[category]] <- as.factor(as.integer(df[[category]]))
  }
  
  return(df)
}

# Apply the function to your dataframe
df <- extract_binary_indicators(df)
df$Match.ID.18Char <- NULL
df$Little.ID <- NULL
df$Big.ID <- NULL
df$Big..Military <- NULL
df$Big.Employer <- NULL
df$Closure.Details <- NULL
df$Big.Open.to.Cross.Gender.Match <- NULL
df$Big.Contact..Interest.Finder...Sports <- NULL
df$Big.Contact..Interest.Finder...Places.To.Go <- NULL
df$Big.Contact..Interest.Finder...Hobbies <- NULL
df$Big.Contact..Interest.Finder...Entertainment <- NULL
df$Big.Contact..Interest.Finder...Hobbies <- NULL
df$Big.Contact..Created.Date <- NULL
df$Big.Enrollment..Created.Date <- NULL
df$Little.Contact..Interest.Finder...Sports <- NULL
df$Little.Contact..Interest.Finder...Outdoors <- NULL
df$Little.Contact..Interest.Finder...Arts <- NULL
df$Little.Contact..Interest.Finder...Places.To.Go <- NULL
df$Little.Contact..Interest.Finder...Hobbies <- NULL
df$Little.Contact..Interest.Finder...Entertainment <- NULL
df$Little.Contact..Interest.Finder...Other.Interests <- NULL
df$Little.Other.Interests <- NULL
df$Little.Contact..Interest.Finder...Career <- NULL
df$Little.Contact..Interest.Finder...Personality <- NULL
df$Little.Contact..Interest.Finder...Three.Wishes <- NULL
df$Little.Other.Interests <- NULL
df$Rationale.for.Match <- NULL
df$Big.County[df$Big.County == ""] <- NA
df$Match.Activation.Date <- as.Date(df$Match.Activation.Date, format="%Y-%m-%d")
df$Big.Approved.Date <- as.Date(df$Big.Approved.Date, format="%Y-%m-%d") 
df$Big.Acceptance.Date <- as.Date(df$Big.Acceptance.Date, format="%Y-%m-%d") 
df$Match.Closure.Meeting.Date <- as.Date(df$Match.Closure.Meeting.Date, format="%Y-%m-%d") 
df$Big.Birthdate <- as.Date(df$Big.Birthdate, format="%Y-%m-%d") 
df$Little.Birthdate <- as.Date(df$Little.Birthdate, format="%Y-%m-%d") 
df$Little.Interview.Date <- as.Date(df$Little.Interview.Date, format="%Y-%m-%d") 
#Function to check if Big and Little ethnicities share any keywords
check_ethnicity_match <- function(df) {
  # Create a new column to store the matching result
  df$Ethnicity_Match <- FALSE
  
  # Loop through each row
  for (i in 1:nrow(df)) {
    # Get the Big and Little race/ethnicity values
    big_race <- df$Big.Race.Ethnicity[i]
    little_race <- df$Little.Participant..Race.Ethnicity[i]
    
    # Skip if either value is NA
    if (is.na(big_race) || is.na(little_race)) {
      df$Ethnicity_Match[i] <- NA
      next
    }
    
    # Convert to character (in case they're factors)
    big_race <- as.character(big_race)
    little_race <- as.character(little_race)
    
    # Split strings by semicolons to handle multiple ethnicities
    big_races <- unlist(strsplit(big_race, ";"))
    little_races <- unlist(strsplit(little_race, ";"))
    
    # Clean up any leading/trailing spaces
    big_races <- trimws(big_races)
    little_races <- trimws(little_races)
    
    # Check if there's any match
    match_found <- FALSE
    for (b in big_races) {
      for (l in little_races) {
        # Extract keywords to compare (simplify the comparison)
        keywords <- c("White", "Black", "Asian", "Hispanic", "Indian", "Alaska", 
                     "Middle Eastern", "North African", "Other")
        
        # Check for each keyword
        for (keyword in keywords) {
          if (grepl(keyword, b, ignore.case = TRUE) && 
              grepl(keyword, l, ignore.case = TRUE)) {
            match_found <- TRUE
            break
          }
        }
        if (match_found) break
      }
      if (match_found) break
    }
    
    # Assign the result
    df$Ethnicity_Match[i] <- match_found
  }
  
  return(df)
}

df <- check_ethnicity_match(df)
df$Big.Race.Ethnicity <- NULL
df$Little.Participant..Race.Ethnicity <- NULL
df$Stage <- factor(ifelse(df$Stage == "Closed", "Closed", "Active"))
df[df == ""] <- NA
df$Big.Languages[df$Big.Languages == ""] <- NA
df$Big.Gender <- factor(df$Big.Gender, 
                        levels = c("Female", "Male"),
                        labels = c("Female", "Male"))

df$Program <- as.factor(df$Program)
df$Program.Type <- as.factor(df$Program.Type)
df$Big.Level.of.Education <- NULL
df$Big.Languages <- NULL
df$Big.Car.Access <- NULL
df$Big.Contact..Preferred.Communication.Type <- NULL
df$Big.Contact..Former.Big.Little <- NULL
df$Big.Contact..Volunteer.Availability <- NULL
df$Little.RTBM.Date.in.MF <- NULL
df$Little.Contact..Language.s..Spoken <- NULL
df$Little.Acceptance.Date <- NULL
df$Little.Application.Received <- NULL
df$Little.Moved.to.RTBM.in.MF <- NULL
df$Little.Mailing.Address.Census.Block.Group <- NULL
df$Little.Acceptance.Date <- NULL
df$Big.Home.Census.Block.Group <- NULL
df$Big.Employer.School.Census.Block.Group <- NULL
df$Little.Gender <- NULL
df$Little.Birthdate <- NULL
df$Little.RTBM.in.Matchforce <- NULL
df$Little.Interview.Date <- NULL
df$Big.Acceptance.Date <- NULL
df$Big.Assessment.Uploaded <- NULL
df$Big.Days.Interview.to.Match <- NULL
df$Big.Days.Interview.to.Acceptance <- NULL
consolidate_counties <- function(county_data, min_frequency = 50) {
  consolidated <- county_data
  county_counts <- table(county_data[county_data != ""])
  rare_counties <- names(county_counts[county_counts < min_frequency])
  consolidated[consolidated %in% rare_counties] <- "Other"
  # Convert to factor with meaningful levels
  consolidated <- factor(consolidated)
  
  return(consolidated)
}

df$County_Factor <- consolidate_counties(df$Big.County)
summary(df$County_Factor)
df$Big.County <- NULL
# Function to categorize text fields based on keywords
categorize_text <- function(text_vector, category_rules, default_category = "Other") {
  result <- rep(default_category, length(text_vector))
  
  if (any(is.na(text_vector))) {
    result[is.na(text_vector)] <- NA
  }
  
  text_vector <- tolower(trimws(text_vector))
  
  for (category_name in names(category_rules)) {
    keywords <- category_rules[[category_name]]
    
    # Check if any keyword appears in each entry
    match_indices <- sapply(text_vector, function(text) any(grepl(paste(keywords, collapse = "|"), text, ignore.case = TRUE)))
    
    # Assign the category where matches occur
    result[match_indices] <- category_name
  }
  
  return(factor(result))
}

# Define category rules for each text field
closure_reason_rules <- list(
  "Scheduling_Issues" = c("schedule", "time", "availability", "busy", "time constraint"),
  "Relationship_Problems" = c("relationship", "conflict", "disagree", "personal", "not compatible", "incompatible", "lost contact", "lost interest"),
  "Relocation" = c("move", "moved", "relocation", "relocate", "different city", "different state"),
  "Family_Issues" = c("family", "parent", "guardian", "parental"),
  "School_Issues" = c("school", "academic", "education", "grade", "graduated", "graduate"),
  "Health_Issues" = c("health", "illness", "medical", "sick", "disease", "covid", "deceased"),
  "Behavior_Issues" = c("behavior", "conduct", "attitude", "disciplin"),
  "Program_Requirements" = c("requirement", "qualify", "eligibility", "criteria", "guideline", "infraction", "expectations", "challenges"),
  "Success" = c("success", "successful")
)

occupation_rules <- list(
  "Business_Finance" = c("account", "financ", "budget", "analyst", "bank", "economic", "market", "business", "consultant", "insurance", "entrepreneur"),
  "Education" = c("teach", "professor", "instructor", "education", "academic", "school", "college", "university"),
  "Healthcare" = c("doctor", "nurse", "medical", "health", "dental", "therapist", "clinic", "hospital", "coach"),
  "Technology" = c("software", "developer", "engineer", "IT", "computer", "tech", "program", "web", "data"),
  "Legal" = c("lawyer", "attorney", "legal", "law", "judge", "paralegal"),
  "Arts_Media" = c("artist", "design", "writer", "media", "journalist", "creative", "music", "film", "arts"),
  "Service_Industry" = c("retail", "sales", "service", "hospitality", "restaurant", "customer", "child"),
  "Trades_Labor" = c("construct", "mechanic", "carpenter", "electric", "plumb", "repair", "builder", "labor"),
  "Student" = c("student", "graduate", "undergrad"),
  "Unknown" = c("unknown"),
  "Retired" = c("retire")
)

df$Closure_Reason_Category <- categorize_text(df$Closure.Reason, closure_reason_rules)
df$Occupation_Category <- categorize_text(df$Big.Occupation, occupation_rules)
summary(df$Closure_Reason_Category)
summary(df$Occupation_Category)
df$Closure.Reason <- NULL
df$Big.Occupation <- NULL
df$Big.Days.Acceptance.to.Match <- abs(df$Big.Days.Acceptance.to.Match)

# Sort the original DataFrame in place
df <- df[order(df$Match.Activation.Date), ]
# Create a factor variable with two levels
df$Big.Enrollment..Record.Type <- factor(
  ifelse(df$Big.Enrollment..Record.Type == "CB Volunteer Enrollment", 
         "CB Volunteer Enrollment", 
         "Others")
)
# Create new categorical variable from Big.Contact..Marital.Status
df$Big.Contact..Marital.Status <- factor(
  case_when(
    df$Big.Contact..Marital.Status == "Single" ~ "Single",
    !is.na(df$Big.Contact..Marital.Status) ~ "Not Single",
    TRUE ~ NA_character_
  ),
  levels = c("Single", "Not Single")
)
df$Stage <- ifelse(df$Stage == "Closed", 1, 0)
```

## Understand and analyze the response variable distributions: Match Length:

```{r}
summary(df$Match.Length)
# Histogram to visualize distribution
hist(df$Match.Length, 
     main="Distribution of Match Length", 
     xlab="Match Length (months)", 
     col="lightblue", 
     breaks=20)

# Density plot
plot(density(df$Match.Length), 
     main="Density Plot of Match Length", 
     xlab="Match Length (months)")

# Boxplot
boxplot(df$Match.Length, 
        main="Boxplot of Match Length", 
        ylab="Match Length (months)")

stats <- summary(df$Match.Length)
mean_value <- mean(df$Match.Length)
q1 <- stats["1st Qu."]
median_value <- stats["Median"]
q3 <- stats["3rd Qu."]

text(x = 1.3, y = mean_value, labels = paste("Mean:", round(mean_value, 2)), col = "black")
text(x = 1.3, y = q1, labels = paste("1Q:", round(q1, 2)), col = "black")
text(x = 1.3, y = median_value, labels = paste("Median:", round(median_value, 2)), col = "black")
text(x = 1.3, y = q3, labels = paste("3Q:", round(q3, 2)), col = "black")

# Check for normality
qqnorm(df$Match.Length)
qqline(df$Match.Length) # Looks exponentially distributed
```
Definitely not normally distributed as expected. Maybe Log transform? Survival analysis?

```{r}
table(df$Closure_Reason_Category)
```
Only 95 defined as successful.

## How do the response variable distributions vary across Program Type?

```{r}
table(df$Program.Type)
df$Program.Type[df$Program.Type == "Site Based Plus"] <- NA # too little records to consider
df$Program.Type <- droplevels(df$Program.Type)
df_filtered <- df %>% filter(!is.na(Program.Type))
# Summary statistics by Program Type
summary_stats <- df %>%
  group_by(Program.Type) %>%
  summarise(
    Mean = mean(Match.Length, na.rm = TRUE),
    Median = median(Match.Length, na.rm = TRUE),
    SD = sd(Match.Length, na.rm = TRUE),
    Q1 = quantile(Match.Length, 0.25, na.rm = TRUE),
    Q3 = quantile(Match.Length, 0.75, na.rm = TRUE)
  )
print(summary_stats)

# Boxplot to visualize distribution of Match.Length by Program.Type
ggplot(df_filtered, aes(x = Program.Type, y = Match.Length, fill = Program.Type)) +
  geom_boxplot(na.rm = TRUE) +
  labs(
    title = "Distribution of Match Length by Program Type",
    x = "Program Type",
    y = "Match Length (months)"
  ) +
  theme_bw()

# Density plot to compare distributions
ggplot(df_filtered, aes(x = Match.Length, fill = Program.Type)) +
  geom_density(alpha = 0.5, na.rm = TRUE) +
  labs(
    title = "Density Plot of Match Length by Program Type",
    x = "Match Length (months)",
    y = "Density"
  ) +
  theme_bw()

# Violin plot for a more detailed view
ggplot(df_filtered, aes(x = Program.Type, y = Match.Length, fill = Program.Type)) +
  geom_violin(na.rm = TRUE) +
  labs(
    title = "Violin Plot of Match Length by Program Type",
    x = "Program Type",
    y = "Match Length (months)"
  ) +
  theme_bw()
```

```{r}
# ANOVA test
anova_result <- aov(Match.Length ~ Program.Type, data = df)
summary(anova_result)

# Chi-Square Test of Independence for Closure Reason
chi_square_result <- chisq.test(table(df$Closure_Reason_Category, df$Program.Type))
print(chi_square_result)

# Kruskal-Wallis test (non-parametric alternative)
kruskal.test(Match.Length ~ Program.Type, data = df)
```

Program.Type a significant predictor of Match Length and Closure Reason

## Response distributions over time

```{r}
ggplot(df, aes(x = Match.Activation.Date, y = Match.Length)) +
  geom_point(alpha = 0.5, color = "blue") +  # Scatter plot with transparency
  geom_smooth(method = "loess", color = "red", se = TRUE) +  # Trend line
  labs(title = "Match Length Over Activation Date",
       x = "Activation Date",
       y = "Match Length") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Clearly older matches have an advantage - look into survival analysis
and Cox potential hazards model.

```{r}
ggplot(df, aes(x = Match.Activation.Date, fill = Closure_Reason_Category)) +
  geom_histogram(position = "stack", bins = 30) +
  labs(
    title = "Closure Reason Categories by Match Activation Date",
    x = "Match Activation Date",
    y = "Count",
    fill = "Closure Reason"
  ) +
  scale_fill_brewer(palette = "Set2") +
  theme_bw() +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

df_monthly <- df %>%
  mutate(Month = floor_date(Match.Activation.Date, "month")) %>%
  group_by(Month, Closure_Reason_Category) %>%
  summarise(Count = n(), .groups = "drop") %>%
  group_by(Month) %>%
  mutate(Proportion = Count / sum(Count))

ggplot(df_monthly, aes(x = Month, y = Proportion, fill = Closure_Reason_Category)) +
  geom_area() +
  labs(
    title = "Monthly Proportion of Closure Reasons",
    x = "Match Activation Date (Monthly)",
    y = "Proportion",
    fill = "Closure Reason"
  ) +
  scale_fill_brewer(palette = "Set2") +
  theme_bw() +
  theme(legend.position = "bottom")
```

Seems to stay relatively consistent - with a rise in 'NA' values in the
bottom due to less match closures.

```{r}
table(df$Stage)
```

2486 match closures, 789 active matches

```{r}
str(df)
```

## What influence do the various Big and/or Little demographic variables have on the response variable distributions?

```{r}
# Analysis of Big Age vs Match Length
ggplot(df, aes(x = Big.Age, y = Match.Length, color = factor(Stage, labels = c("Active", "Closed")))) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess", color = "blue") +
  labs(
    title = "Match Length by Big's Age",
    x = "Big's Age",
    y = "Match Length (months)",
    color = "Stage"  # Legend title
  ) +
  theme_bw()

# Age group analysis
df$Age_Group <- cut(df$Big.Age, 
                    breaks = c(0, 25, 35, 45, 55, 65, 100),
                    labels = c("18-25", "26-35", "36-45", "46-55", "56-65", "65+"))

age_summary <- df %>%
  group_by(Age_Group) %>%
  summarise(
    Mean_Length = mean(Match.Length, na.rm = TRUE),
    Median_Length = median(Match.Length, na.rm = TRUE),
    Count = n()
  )
age_summary
```

```{r}
# Big Gender analysis
gender_summary <- df %>%
  group_by(Big.Gender) %>%
  summarise(
    Mean_Length = mean(Match.Length, na.rm = TRUE),
    Median_Length = median(Match.Length, na.rm = TRUE),
    Count = n()
  )
gender_summary
# Box plot of match length by Big's gender
ggplot(df, aes(x = Big.Gender, y = Match.Length, fill = Big.Gender)) +
  geom_boxplot() +
  labs(title = "Match Length by Big's Gender",
       x = "Gender",
       y = "Match Length (months)") +
  theme_bw()

# Statistical test for gender difference
t.test(Match.Length ~ Big.Gender, data = df)
```
Statistically discernable difference in match length and gender. Longer for male bigs.

```{r}
# Ethnicity match analysis
ethnicity_summary <- df %>%
  group_by(Ethnicity_Match) %>%
  summarise(
    Mean_Length = mean(Match.Length, na.rm = TRUE),
    Median_Length = median(Match.Length, na.rm = TRUE),
    Count = n()
  )
ethnicity_summary
# Box plot for ethnicity match
ggplot(df %>% filter(!is.na(Ethnicity_Match)), 
       aes(x = factor(Ethnicity_Match), y = Match.Length, fill = factor(Ethnicity_Match))) +
  geom_boxplot() +
  labs(title = "Match Length by Ethnicity Match",
       x = "Ethnicity Match",
       y = "Match Length (months)") +
  scale_x_discrete(labels = c("FALSE" = "No Match", "TRUE" = "Match")) +
  scale_fill_discrete(name = "Ethnicity Match", labels = c("No Match", "Match")) +
  theme_bw()
```
Statsitically discerinible difference for match length and ethnicity match - longer if same ethnicity.

```{r}
# Interest and proximity analysis
interest_summary <- df %>%
  group_by(has_interests) %>%
  summarise(
    Mean_Length = mean(Match.Length, na.rm = TRUE),
    Median_Length = median(Match.Length, na.rm = TRUE),
    Count = n()
  )
interest_summary
proximity_summary <- df %>%
  group_by(has_proximity) %>%
  summarise(
    Mean_Length = mean(Match.Length, na.rm = TRUE),
    Median_Length = median(Match.Length, na.rm = TRUE),
    Count = n()
  )
proximity_summary
```
Statistically discernible difference for close distance - longer for closer.

```{r}
# County analysis
county_summary <- df %>%
  group_by(County_Factor) %>%
  summarise(
    Mean_Length = mean(Match.Length, na.rm = TRUE),
    Median_Length = median(Match.Length, na.rm = TRUE),
    Count = n()
  ) %>%
  arrange(desc(Mean_Length))
county_summary
# Boxplot for counties
ggplot(df, aes(x = reorder(County_Factor, Match.Length, FUN = median), 
               y = Match.Length, fill = County_Factor)) +
  geom_boxplot() +
  labs(title = "Match Length by County",
       x = "County",
       y = "Match Length (months)") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")
```


```{r}
# Occupation analysis
occupation_summary <- df %>%
  group_by(Occupation_Category) %>%
  summarise(
    Mean_Length = mean(Match.Length, na.rm = TRUE),
    Median_Length = median(Match.Length, na.rm = TRUE),
    Count = n()
  ) %>%
  arrange(desc(Mean_Length))
occupation_summary

ggplot(df, aes(x = reorder(Occupation_Category, Match.Length, FUN = median), 
               y = Match.Length, fill = Occupation_Category)) +
  geom_boxplot() +
  labs(title = "Match Length by Big's Occupation",
       x = "Occupation Category",
       y = "Match Length (months)") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")

```
Students on average have shorter matches.


```{r}
df_marital <- df %>% filter(!is.na(Big.Contact..Marital.Status))
marital_summary <- df_marital %>%
  group_by(Big.Contact..Marital.Status) %>%
  summarise(
    Mean_Length = mean(Match.Length, na.rm = TRUE),
    Median_Length = median(Match.Length, na.rm = TRUE),
    Count = n()
  )
marital_summary
```
Of those which are available, not single bigs have longer match lengths - more stability?

## Analysis of closure reasons by demographics
```{r}
age_closure <- df %>%
  filter(!is.na(Age_Group), !is.na(Closure_Reason_Category)) %>%
  group_by(Age_Group, Closure_Reason_Category) %>%
  summarise(Count = n(), .groups = "drop") %>%
  group_by(Age_Group) %>%
  mutate(Percentage = Count / sum(Count) * 100)

gender_closure <- df %>%
  filter(!is.na(Big.Gender), !is.na(Closure_Reason_Category)) %>%
  group_by(Big.Gender, Closure_Reason_Category) %>%
  summarise(Count = n(), .groups = "drop") %>%
  group_by(Big.Gender) %>%
  mutate(Percentage = Count / sum(Count) * 100)

# Visualize closure reasons by gender
ggplot(gender_closure, aes(x = Big.Gender, y = Percentage, fill = Closure_Reason_Category)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(title = "Closure Reasons by Big's Gender",
       x = "Gender",
       y = "Percentage",
       fill = "Closure Reason") +
  theme_bw() +
  theme(legend.position = "right")
```
Roughly equal - more family issues for males? more health issues for females?

```{r}
# Statistical tests for demographic effects on match length
# ANOVA for County effect
county_anova <- aov(Match.Length ~ County_Factor, data = df)
summary(county_anova)

# ANOVA for Occupation effect
occupation_anova <- aov(Match.Length ~ Occupation_Category, data = df)
summary(occupation_anova)

# ANOVA for Age Group effect
age_anova <- aov(Match.Length ~ Age_Group, data = df)
summary(age_anova)

# Simple regression model
model <- lm(Match.Length ~ Big.Age + Big.Gender + Program.Type + Ethnicity_Match + 
             has_interests + has_proximity + County_Factor + Occupation_Category, 
           data = df %>% filter(!is.na(Ethnicity_Match)))
summary(model)
```
Very poor predictive performance. But many important predictors.


## Looking at interests
```{r}
df_with_indicators <- df
summary_indicators <- df_with_indicators %>%
  summarise(across(c(has_interests, personality_compatibility, has_proximity, 
                    has_commitment, has_experience, has_preference,
                    has_challenges, has_goals), 
                   ~sum(as.integer(as.character(.)) == 1, na.rm = TRUE)))
summary_indicators

# Calculate correlation with match length
indicator_correlations <- df_with_indicators %>%
  select(Match.Length, has_interests, personality_compatibility, has_proximity, 
         has_commitment, has_experience, has_preference,
         has_challenges, has_goals) %>%
  mutate(across(has_interests:has_goals, ~as.numeric(as.character(.)))) %>%
  cor(use = "pairwise.complete.obs")

print(indicator_correlations["Match.Length", ])

# Visualize the distribution of match length by each indicator
library(gridExtra)
```

```{r, fig.width=8, fig.height=12}
# Function to create box plots
create_boxplot <- function(df, var) {
  ggplot(df, aes_string(x = var, y = "Match.Length", fill = var)) +
    geom_boxplot() +
    labs(title = paste("Match Length by", var),
         x = var,
         y = "Match Length (months)") +
    theme_bw() +
    theme(legend.position = "none") +
    scale_x_discrete(labels = c("0" = "No", "1" = "Yes"))
}

# Create a list of plots
plot_list <- lapply(c("has_interests", "personality_compatibility", "has_proximity", 
                      "has_commitment", "has_experience", "has_preference",
                      "has_challenges", "has_goals"), 
                    function(var) create_boxplot(df_with_indicators, var))

# Arrange plots in a grid
grid.arrange(grobs = plot_list, ncol = 2)
```
has_experience not statistically discernible and present in data much.

```{r, fig.width=4, fig.height=4}
# Analyze the impact of different interest combinations
df_with_indicators$interest_count <- rowSums(sapply(df_with_indicators[, c("has_interests", 
                                                                         "personality_compatibility", 
                                                                         "has_proximity", 
                                                                         "has_commitment", 
                                                                         "has_experience", 
                                                                         "has_preference",
                                                                         "has_challenges", 
                                                                         "has_goals")], 
                                                  function(x) as.integer(as.character(x))))

# Analyze relationship between number of interest categories and match length
interest_count_summary <- df_with_indicators %>%
  group_by(interest_count) %>%
  summarise(
    Mean_Length = mean(Match.Length, na.rm = TRUE),
    Median_Length = median(Match.Length, na.rm = TRUE),
    Count = n()
  )

print(interest_count_summary)

# Visualize relationship between interest count and match length
ggplot(df_with_indicators, aes(x = factor(interest_count), y = Match.Length)) +
  geom_boxplot(fill = "lightblue") +
  labs(title = "Match Length by Number of Interest Categories Mentioned",
       x = "Number of Interest Categories",
       y = "Match Length (months)") +
  theme_bw()

# Test statistical significance
interest_model <- lm(Match.Length ~ has_interests + personality_compatibility + has_proximity + 
                       has_commitment + has_experience + has_preference +
                       has_challenges + has_goals, data = df_with_indicators)
summary(interest_model)
```

## Successful match?
```{r}
# A match is successful if:
# 1. It's active 
# 2. It has a long duration 
# 3. Closure reason is "Success"

df$successful_match <- FALSE

# Active matches (no closure date)
df$successful_match[df$Stage == 0] <- TRUE
sum(df$successful_match[df$Stage == 0])

# Matches with "Success" as closure reason
df$successful_match[df$Closure_Reason_Category == "Success"] <- TRUE

# Long duration matches (define long as above the 80th percentile)
long_duration_threshold <- quantile(df$Match.Length, 0.8, na.rm = TRUE)
df$successful_match[df$Match.Length > long_duration_threshold & 
                      !is.na(df$Match.Length)] <- TRUE

# Success rate
success_rate <- mean(df$successful_match, na.rm = TRUE)
print(paste("Overall success rate:", round(success_rate * 100, 1), "%"))
```
A third of the matches are successful (maybe slightly biased because many matches just started).

analyzing factors associated with successful matches
```{r}
age_group_success <- aggregate(successful_match ~ Age_Group, data = df, FUN = mean)
age_group_success$count <- aggregate(successful_match ~ Age_Group, data = df, FUN = length)$successful_match
age_group_success <- age_group_success[order(-age_group_success$successful_match),]
print("Success rate by mentor age group:")
print(age_group_success)

gender_success <- aggregate(successful_match ~ Big.Gender, data = df, FUN = mean)
gender_success$count <- aggregate(successful_match ~ Big.Gender, data = df, FUN = length)$successful_match
print("Success rate by mentor gender:")
print(gender_success)

program_type_success <- aggregate(successful_match ~ Program.Type, data = df, FUN = mean)
program_type_success$count <- aggregate(successful_match ~ Program.Type, data = df, FUN = length)$successful_match
print("Success rate by program type:")
print(program_type_success)

ethnicity_match_success <- aggregate(successful_match ~ Ethnicity_Match, data = df, FUN = mean)
ethnicity_match_success$count <- aggregate(successful_match ~ Ethnicity_Match, data = df, FUN = length)$successful_match
print("Success rate by ethnicity match:")
print(ethnicity_match_success)

occupation_success <- aggregate(successful_match ~ Occupation_Category, data = df, FUN = mean)
occupation_success$count <- aggregate(successful_match ~ Occupation_Category, data = df, FUN = length)$successful_match
occupation_success <- occupation_success[order(-occupation_success$successful_match),]
print("Success rate by occupation category:")
print(occupation_success)

county_success <- aggregate(successful_match ~ County_Factor, data = df, FUN = mean)
county_success$count <- aggregate(successful_match ~ County_Factor, data = df, FUN = length)$successful_match
county_success <- county_success[order(-county_success$successful_match),]
print("Success rate by county:")
print(county_success)

# Create a function to check success rate for binary factors
check_binary_factor <- function(factor_name) {
  formula <- as.formula(paste("successful_match ~", factor_name))
  success_rate <- aggregate(formula, data = df, FUN = mean)
  success_rate$count <- aggregate(formula, data = df, FUN = length)$successful_match
  print(paste("Success rate by", factor_name, ":"))
  print(success_rate)
}

compatibility_factors <- c("has_interests", "personality_compatibility", "has_proximity", 
                          "has_commitment", "has_experience", "has_preference",
                          "has_challenges", "has_goals")

for (factor in compatibility_factors) {
  check_binary_factor(factor)
}

# Logistic regression to identify key predictors of success
df$successful_match_numeric <- as.numeric(df$successful_match)

model <- glm(successful_match_numeric ~ Big.Age + Big.Gender + Program.Type + 
              Ethnicity_Match + County_Factor + Occupation_Category + Age_Group +
              has_interests + personality_compatibility + has_proximity + 
              has_commitment + has_experience + has_preference + has_challenges + has_goals,
            family = binomial(link = "logit"), data = df)

summary_model <- summary(model)
print("Logistic regression results (key predictors of successful matches):")
print(summary_model)

significant_predictors <- summary_model$coefficients[summary_model$coefficients[,4] < 0.05,]
print("Significant predictors of match success:")
print(significant_predictors)
```

## Logistic Regresison on whether Match was a Success
```{r}
filtered_data <- df %>%
  filter(Match.Activation.Date < as.Date("2023-11-01"))
set.seed(8080)
sample_size <- floor(0.8 * nrow(filtered_data))
train_indices <- sample(seq_len(nrow(filtered_data)), size = sample_size)
train_data <- filtered_data[train_indices, ]
test_data <- filtered_data[-train_indices, ]

logistic_model_train <- glm(successful_match ~ Big.Age + Big.Gender + Program.Type + 
                            has_interests + personality_compatibility + 
                            has_proximity + has_commitment + has_experience + 
                            has_preference + has_challenges + has_goals + 
                            Ethnicity_Match + County_Factor + Closure_Reason_Category + 
                            Occupation_Category + Age_Group, 
                            data = train_data, 
                            family = binomial(link = "logit"))


predictions <- predict(logistic_model_train, newdata = test_data, type = "response")
predicted_classes <- ifelse(predictions > 0.5, 1, 0)

# Evaluate the model (e.g., using confusion matrix)
confusion_matrix <- table(Predicted = predicted_classes, Actual = test_data$successful_match)
print(confusion_matrix)

# Calculate accuracy
accuracy <- sum(diag(confusion_matrix)) / sum(confusion_matrix)
print(paste("Accuracy:", accuracy))
```
```{r}
summary(logistic_model_train)
```
```{r}
library(car)   
vif(logistic_model_train)
```

```{r}
boxTidwell(successful_match ~ Big.Age + Match.Length, data = df_filtered)
```
```{r}
plot(logistic_model_train, which = 4) 
```
S` 
'b 

## Survival analysis on successful matches
```{r}
library(survival)
cox_model <- coxph(Surv(Match.Length, successful_match_numeric) ~ Big.Gender + Big.Age + Program.Type + Occupation_Category + has_interests + has_proximity + has_goals + Ethnicity_Match , data = df)
summary(cox_model)
```
Surprisingly very good model. Program type, big age and shared interest seem to be the most telling signs of match length by succesful match.

```{r}
cox_test <- cox.zph(cox_model)
print(cox_test)
```

```{r}
# distinct time intervals based on data

early_matches <- subset(df, Match.Length <= 6)  # 0-6 months
medium_matches <- subset(df, Match.Length > 6 & Match.Length <= 12)  # 6-12 months
long_matches <- subset(df, Match.Length > 12)  # > 12 months

early_model <- coxph(
  Surv(Match.Length, successful_match_numeric) ~ 
    Big.Gender + Big.Age + Program.Type + 
    Occupation_Category + has_interests + has_proximity + 
    has_goals + Ethnicity_Match, 
  data = early_matches
)

medium_model <- coxph(
  Surv(Match.Length, successful_match_numeric) ~ 
    Big.Gender + Big.Age + Program.Type + 
    Occupation_Category + has_interests + has_proximity + 
    has_goals + Ethnicity_Match, 
  data = medium_matches
)

long_model <- coxph(
  Surv(Match.Length, successful_match_numeric) ~ 
    Big.Gender + Big.Age + Program.Type + 
    Occupation_Category + has_interests + has_proximity + 
    has_goals + Ethnicity_Match, 
  data = long_matches
)

cox_test_early <- cox.zph(early_model)
cox_test_medium <- cox.zph(medium_model)
cox_test_long <- cox.zph(long_model)

summary(early_model)
summary(medium_model)
summary(long_model)
```

Early:
Program.TypeSite Based Facilitated, has interest
Concordance = 0.703

Middle:
Health care, interests, goals
Concordance = 0.781

Late:
Big.Age, Program.Typesite, has interests
Concordance = 0.627

Shared Interests have a dramatic effect on match success across all time periods, but with varying magnitude:

Short-term: HR = 12.11
Medium-term: HR = 43.01
Long-term: HR = 2.12
This suggests interests are absolutely critical for early and medium success but less deterministic for long-term outcomes


Program Types show a pattern:
Site Based Facilitated programs excel in short-term (HR = 2.32) and long-term (HR = 3.99)
Site programs perform poorly short-term (HR = 0.39) but succeed long-term (HR = 2.26)
This suggests different program structures have different timelines for success

Goals matter most in medium-term matches (HR = 1.81)
Mentor Age becomes significant only in long-term matches

```{r}
cox_test <- cox.zph(early_model)
print(cox_test)
cox_test <- cox.zph(medium_model)
print(cox_test)
cox_test <- cox.zph(long_model)
print(cox_test)
```
Early model valid, others not so.

```{r}
df_filtered <- df[df$Match.Length > 0, ]
weibull_aft <- survreg(
  Surv(Match.Length, successful_match_numeric) ~ 
    Big.Gender + Big.Age + Program.Type + 
    Occupation_Category + has_interests + has_proximity + 
    has_goals + Ethnicity_Match,
  data = df_filtered, 
  dist = "weibull"
)
summary(weibull_aft)
```

positive coefficients indicate longer survival times, which in this case means longer-lasting matches. The scale parameter being less than 1 (0.52) suggests that the hazard rate (risk of match ending) decreases over time, meaning matches that survive the initial period tend to become more stable.

```{r}
# Load necessary packages
library(survival)
library(survminer)
library(car)
library(ggplot2)
library(flexsurv)


# Weibull distribution appropriateness

# Extract model residuals
resids <- residuals(weibull_aft)

# QQ plot of residuals
qqnorm(resids)
qqline(resids)

# Create a log(-log(survival)) vs log(time) plot to check linearity
# First, create a simple Weibull model with just intercept for basic check
simple_model <- survreg(Surv(Match.Length, successful_match_numeric) ~ 1, 
                        data = df_filtered, dist = "weibull")

# Generate survival estimates
surv_obj <- Surv(df_filtered$Match.Length, df_filtered$successful_match_numeric)
surv_fit <- survfit(surv_obj ~ 1)

# Extract survival estimates and times
surv_data <- data.frame(
  time = surv_fit$time,
  surv = surv_fit$surv
)

# Calculate log(-log(survival)) and log(time)
surv_data$log_time <- log(surv_data$time)
surv_data$log_neg_log_surv <- log(-log(surv_data$surv))

# Plot (should be approximately linear for Weibull)
ggplot(surv_data, aes(x = log_time, y = log_neg_log_surv)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "log(time)", y = "log(-log(survival))",
       title = "Checking Weibull assumption: log(-log(S)) vs log(t)") +
  theme_minimal()

# Compare with other distributions using AIC
model_weibull <- survreg(formula = Surv(Match.Length, successful_match_numeric) ~ 
    Big.Gender + Big.Age + Program.Type + Occupation_Category + 
    has_interests + has_proximity + has_goals + Ethnicity_Match, 
    data = df_filtered, dist = "weibull")

model_exponential <- survreg(formula = Surv(Match.Length, successful_match_numeric) ~ 
    Big.Gender + Big.Age + Program.Type + Occupation_Category + 
    has_interests + has_proximity + has_goals + Ethnicity_Match, 
    data = df_filtered, dist = "exponential")

model_lognormal <- survreg(formula = Surv(Match.Length, successful_match_numeric) ~ 
    Big.Gender + Big.Age + Program.Type + Occupation_Category + 
    has_interests + has_proximity + has_goals + Ethnicity_Match, 
    data = df_filtered, dist = "lognormal")

model_loglogistic <- survreg(formula = Surv(Match.Length, successful_match_numeric) ~ 
    Big.Gender + Big.Age + Program.Type + Occupation_Category + 
    has_interests + has_proximity + has_goals + Ethnicity_Match, 
    data = df_filtered, dist = "loglogistic")

# Compare AIC values
aic_values <- c(
  Weibull = AIC(model_weibull),
  Exponential = AIC(model_exponential),
  Lognormal = AIC(model_lognormal),
  Loglogistic = AIC(model_loglogistic)
)
print("AIC comparison of different distributions:")
print(aic_values)
print(paste("Best model based on AIC:", names(which.min(aic_values))))


# Check proportional hazards

# Convert survreg model to Cox model for diagnostics
cox_model <- coxph(Surv(Match.Length, successful_match_numeric) ~ 
    Big.Gender + Big.Age + Program.Type + Occupation_Category + 
    has_interests + has_proximity + has_goals + Ethnicity_Match, 
    data = df_filtered)

# Test proportional hazards assumption
ph_test <- cox.zph(cox_model)
print(ph_test)

# Plot Schoenfeld residuals
par(mfrow = c(3, 3))
plot(ph_test)
par(mfrow = c(1, 1))


# Check for influential observations

# Calculate deviance residuals
dev_resids <- residuals(model, type = "deviance")

# Plot deviance residuals vs fitted values
plot(predict(model), dev_resids,
     xlab = "Fitted Values", ylab = "Deviance Residuals",
     main = "Residual Plot")
abline(h = 0, lty = 2)

# Create index plot for deviance residuals
plot(dev_resids, type = "h",
     xlab = "Observation Index", ylab = "Deviance Residuals",
     main = "Index Plot of Deviance Residuals")
abline(h = c(-3, 0, 3), lty = c(2, 1, 2))

```

