---
title: "MinneMUDAC DS - Novice Questions"
subtitle: "Noah Lee"
output:
  pdf_document: default
  html_document: default
editor_options: 
  markdown: 
    wrap: 72
---

```{r}
#| include: false
library(tidyverse)
library(tidymodels)
library(naniar)    
library(dplyr)
library(ggformula)
library(ggplot2) 
library(GGally)

library(lubridate)
library(ranger)   
library(workflows) 
library(recipes) 
tidymodels_prefer(quiet = TRUE) 
```

Preprocessing:

```{r}
df <- read.csv('../Data/Novice.csv')
extract_binary_indicators <- function(df) {
  # Initialize new columns with FALSE (0)
  interest_categories <- c("has_interests", "personality_compatibility", "has_proximity", 
                          "has_commitment", "has_experience", "has_preference",
                          "has_challenges", "has_goals")
  
  for (category in interest_categories) {
    df[[category]] <- FALSE  # Initialize with FALSE for all rows
  }
  
  # Define keywords for each category
  keywords <- list(
    has_interests = c("outdoors", "hiking", "biking", "fishing", "camping", "parks", "nature", "gardening", "swimming", "sledding", "horseback riding", "skateboarding", "snowboarding", "ice skating", "picnics", "planting", "rock climbing", "feeding ducks", "flying kites", "sports", "basketball", "football", "soccer", "baseball", "hockey", "bowling", "tennis", "running", "yoga", "Zumba", "gymnastics", "arts", "crafts", "drawing", "painting", "pottery", "sewing", "knitting", "photography", "model cars", "model planes", "creative activities", "acting", "singing", "dancing", "playing musical instruments", "writing","indoor activities", "reading", "cooking", "baking", "board games", "video games", "puzzles", "Lego", "animals", "dogs", "cats", "horses", "pets", "animal care", "learning", "science", "math", "history", "social studies", "STEM projects", "other interests", "fashion", "hair", "nails", "volunteering", "museums", "libraries", "zoos", "movies", "plays"),
    personality_compatibility = c("outgoing", "talkative", "bubbly", "energetic", "enthusiastic", "charismatic", 
                                "shy", "reserved", "quiet", "introverted", "soft-spoken", "calm", 
                                "adventurous", "curious", "exploratory", "open to new things", 
                                "friendly", "kind", "sweet", "thoughtful", "empathetic", 
                                "funny", "goofy", "humorous", "light-hearted", 
                                "mature", "respectful", "responsible", "thoughtful", 
                                "active", "sporty", "energetic", "athletic", 
                                "creative", "imaginative", "artistic", "crafty", 
                                "patient", "calm", "steady", "nurturing"),
    has_proximity = c("miles", "minutes", "close", "far", "convenient", "driving", "traffic", "commute"),
    has_commitment = c("long-term", "committed", "consistent", "reliable", "short-term", "temporary", "limited time", "2-4 times a month", "weekly"),
    has_experience = c("child experience", "nanny", "teacher", "coach", "mentor", "social work", "counseling", "teaching", "overcoming challenges", "mental health"),
    has_preference= c("age", "younger", "older", "in 20s", "gender", "male", "female", "religion", "Christian", "Catholic", "cultural background", "African American", "Hispanic", "non-smoker", "no guns"),
    has_challenges = c("behavioral challenges", "ADHD", "unmedicated", "redirection", "mental health", "depression", "anxiety", "PTSD", "family dynamics", "divorce", "strained relationships", "bullying", "picked on", "self-esteem", "academic challenges", "tutoring", "homework help"),
    has_goals = c("self-esteem", "confidence", "self-image", "leadership", "decision-making", "independence", "academic success", "math", "science", "reading", "social skills", "communication", "making friends", "exploration", "trying new things", "learning new skills")
  )
  
  # Check if Rationale.for.Match column exists in the dataframe
  if (!"Rationale.for.Match" %in% names(df)) {
    warning("Column 'Rationale.for.Match' not found in dataframe. No keywords will be extracted.")
    # Return dataframe with all FALSE values
    return(df)
  }
  
  # Process each row
  for (i in 1:nrow(df)) {
    rationale <- df$Rationale.for.Match[i]
    
    # Skip if rationale is NA or empty
    if (is.na(rationale) || rationale == "") {
      next
    }
    
    # Check for keywords in each category
    for (category in names(keywords)) {
      category_keywords <- keywords[[category]]
      for (keyword in category_keywords) {
        if (grepl(keyword, rationale, ignore.case = TRUE)) {
          df[[category]][i] <- TRUE
          break  # Once we find a match, no need to check other keywords in this category
        }
      }
    }
  }
  
  # Convert logical columns to factors (0/1)
  for (category in interest_categories) {
    df[[category]] <- as.factor(as.integer(df[[category]]))
  }
  
  return(df)
}

# Apply the function to your dataframe
df <- extract_binary_indicators(df)
df$Match.ID.18Char <- NULL
df$Little.ID <- NULL
df$Big.ID <- NULL
df$Big..Military <- NULL
df$Big.Employer <- NULL
df$Closure.Details <- NULL
df$Big.Open.to.Cross.Gender.Match <- NULL
df$Big.Contact..Interest.Finder...Sports <- NULL
df$Big.Contact..Interest.Finder...Places.To.Go <- NULL
df$Big.Contact..Interest.Finder...Hobbies <- NULL
df$Big.Contact..Interest.Finder...Entertainment <- NULL
df$Big.Contact..Interest.Finder...Hobbies <- NULL
df$Big.Contact..Created.Date <- NULL
df$Big.Enrollment..Created.Date <- NULL
df$Little.Contact..Interest.Finder...Sports <- NULL
df$Little.Contact..Interest.Finder...Outdoors <- NULL
df$Little.Contact..Interest.Finder...Arts <- NULL
df$Little.Contact..Interest.Finder...Places.To.Go <- NULL
df$Little.Contact..Interest.Finder...Hobbies <- NULL
df$Little.Contact..Interest.Finder...Entertainment <- NULL
df$Little.Contact..Interest.Finder...Other.Interests <- NULL
df$Little.Other.Interests <- NULL
df$Little.Contact..Interest.Finder...Career <- NULL
df$Little.Contact..Interest.Finder...Personality <- NULL
df$Little.Contact..Interest.Finder...Three.Wishes <- NULL
df$Little.Other.Interests <- NULL
df$Rationale.for.Match <- NULL
df$Big.County[df$Big.County == ""] <- NA
df$Match.Activation.Date <- as.Date(df$Match.Activation.Date, format="%Y-%m-%d")
df$Big.Approved.Date <- as.Date(df$Big.Approved.Date, format="%Y-%m-%d") 
df$Big.Acceptance.Date <- as.Date(df$Big.Acceptance.Date, format="%Y-%m-%d") 
df$Match.Closure.Meeting.Date <- as.Date(df$Match.Closure.Meeting.Date, format="%Y-%m-%d") 
df$Big.Birthdate <- as.Date(df$Big.Birthdate, format="%Y-%m-%d") 
df$Little.Birthdate <- as.Date(df$Little.Birthdate, format="%Y-%m-%d") 
df$Little.Interview.Date <- as.Date(df$Little.Interview.Date, format="%Y-%m-%d") 
#Function to check if Big and Little ethnicities share any keywords
check_ethnicity_match <- function(df) {
  # Create a new column to store the matching result
  df$Ethnicity_Match <- FALSE
  
  # Loop through each row
  for (i in 1:nrow(df)) {
    # Get the Big and Little race/ethnicity values
    big_race <- df$Big.Race.Ethnicity[i]
    little_race <- df$Little.Participant..Race.Ethnicity[i]
    
    # Skip if either value is NA
    if (is.na(big_race) || is.na(little_race)) {
      df$Ethnicity_Match[i] <- NA
      next
    }
    
    # Convert to character (in case they're factors)
    big_race <- as.character(big_race)
    little_race <- as.character(little_race)
    
    # Split strings by semicolons to handle multiple ethnicities
    big_races <- unlist(strsplit(big_race, ";"))
    little_races <- unlist(strsplit(little_race, ";"))
    
    # Clean up any leading/trailing spaces
    big_races <- trimws(big_races)
    little_races <- trimws(little_races)
    
    # Check if there's any match
    match_found <- FALSE
    for (b in big_races) {
      for (l in little_races) {
        # Extract keywords to compare (simplify the comparison)
        keywords <- c("White", "Black", "Asian", "Hispanic", "Indian", "Alaska", 
                     "Middle Eastern", "North African", "Other")
        
        # Check for each keyword
        for (keyword in keywords) {
          if (grepl(keyword, b, ignore.case = TRUE) && 
              grepl(keyword, l, ignore.case = TRUE)) {
            match_found <- TRUE
            break
          }
        }
        if (match_found) break
      }
      if (match_found) break
    }
    
    # Assign the result
    df$Ethnicity_Match[i] <- match_found
  }
  
  return(df)
}

df <- check_ethnicity_match(df)
df$Big.Race.Ethnicity <- NULL
df$Little.Participant..Race.Ethnicity <- NULL
df$Stage <- factor(ifelse(df$Stage == "Closed", "Closed", "Active"))
df[df == ""] <- NA
df$Big.Languages[df$Big.Languages == ""] <- NA
df$Big.Gender <- factor(df$Big.Gender, 
                        levels = c("Female", "Male"),
                        labels = c("Female", "Male"))

df$Program <- as.factor(df$Program)
df$Program.Type <- as.factor(df$Program.Type)
df$Big.Level.of.Education <- NULL
df$Big.Languages <- NULL
df$Big.Car.Access <- NULL
df$Big.Contact..Preferred.Communication.Type <- NULL
df$Big.Contact..Former.Big.Little <- NULL
df$Big.Contact..Volunteer.Availability <- NULL
df$Little.RTBM.Date.in.MF <- NULL
df$Little.Contact..Language.s..Spoken <- NULL
df$Little.Acceptance.Date <- NULL
df$Little.Application.Received <- NULL
df$Little.Moved.to.RTBM.in.MF <- NULL
df$Little.Mailing.Address.Census.Block.Group <- NULL
df$Little.Acceptance.Date <- NULL
df$Big.Home.Census.Block.Group <- NULL
df$Big.Employer.School.Census.Block.Group <- NULL
df$Little.Gender <- NULL
df$Little.Birthdate <- NULL
df$Little.RTBM.in.Matchforce <- NULL
df$Little.Interview.Date <- NULL
df$Big.Acceptance.Date <- NULL
df$Big.Assessment.Uploaded <- NULL
df$Big.Days.Interview.to.Match <- NULL
df$Big.Days.Interview.to.Acceptance <- NULL
consolidate_counties <- function(county_data, min_frequency = 50) {
  consolidated <- county_data
  county_counts <- table(county_data[county_data != ""])
  rare_counties <- names(county_counts[county_counts < min_frequency])
  consolidated[consolidated %in% rare_counties] <- "Other"
  # Convert to factor with meaningful levels
  consolidated <- factor(consolidated)
  
  return(consolidated)
}

df$County_Factor <- consolidate_counties(df$Big.County)
summary(df$County_Factor)
df$Big.County <- NULL
# Function to categorize text fields based on keywords
categorize_text <- function(text_vector, category_rules, default_category = "Other") {
  result <- rep(default_category, length(text_vector))
  
  if (any(is.na(text_vector))) {
    result[is.na(text_vector)] <- NA
  }
  
  text_vector <- tolower(trimws(text_vector))
  
  for (category_name in names(category_rules)) {
    keywords <- category_rules[[category_name]]
    
    # Check if any keyword appears in each entry
    match_indices <- sapply(text_vector, function(text) any(grepl(paste(keywords, collapse = "|"), text, ignore.case = TRUE)))
    
    # Assign the category where matches occur
    result[match_indices] <- category_name
  }
  
  return(factor(result))
}

# Define category rules for each text field
closure_reason_rules <- list(
  "Scheduling_Issues" = c("schedule", "time", "availability", "busy", "time constraint"),
  "Relationship_Problems" = c("relationship", "conflict", "disagree", "personal", "not compatible", "incompatible", "lost contact", "lost interest"),
  "Relocation" = c("move", "moved", "relocation", "relocate", "different city", "different state"),
  "Family_Issues" = c("family", "parent", "guardian", "parental"),
  "School_Issues" = c("school", "academic", "education", "grade", "graduated", "graduate"),
  "Health_Issues" = c("health", "illness", "medical", "sick", "disease", "covid", "deceased"),
  "Behavior_Issues" = c("behavior", "conduct", "attitude", "disciplin"),
  "Program_Requirements" = c("requirement", "qualify", "eligibility", "criteria", "guideline", "infraction", "expectations", "challenges"),
  "Success" = c("success", "successful"),
  "Little_Age" = c("age of little", "too young", "not old enough", "underage", "minor", "young age", "small child", "infant", "toddler")
)

occupation_rules <- list(
  "Business_Finance" = c("account", "financ", "budget", "analyst", "bank", "economic", "market", "business", "consultant", "insurance", "entrepreneur"),
  "Education" = c("teach", "professor", "instructor", "education", "academic", "school", "college", "university"),
  "Healthcare" = c("doctor", "nurse", "medical", "health", "dental", "therapist", "clinic", "hospital", "coach"),
  "Technology" = c("software", "developer", "engineer", "IT", "computer", "tech", "program", "web", "data"),
  "Legal" = c("lawyer", "attorney", "legal", "law", "judge", "paralegal"),
  "Arts_Media" = c("artist", "design", "writer", "media", "journalist", "creative", "music", "film", "arts"),
  "Service_Industry" = c("retail", "sales", "service", "hospitality", "restaurant", "customer", "child"),
  "Trades_Labor" = c("construct", "mechanic", "carpenter", "electric", "plumb", "repair", "builder", "labor"),
  "Student" = c("student", "graduate", "undergrad"),
  "Unknown" = c("unknown"),
  "Retired" = c("retire")
)

df$Closure_Reason_Category <- categorize_text(df$Closure.Reason, closure_reason_rules)
df$Occupation_Category <- categorize_text(df$Big.Occupation, occupation_rules)
summary(df$Closure_Reason_Category)
summary(df$Occupation_Category)
df$Closure.Reason <- NULL
df$Big.Occupation <- NULL
df$Big.Days.Acceptance.to.Match <- abs(df$Big.Days.Acceptance.to.Match)

# Sort the original DataFrame in place
df <- df[order(df$Match.Activation.Date), ]
# Create a factor variable with two levels
df$Big.Enrollment..Record.Type <- factor(
  ifelse(df$Big.Enrollment..Record.Type == "CB Volunteer Enrollment", 
         "CB Volunteer Enrollment", 
         "Others")
)
# Create new categorical variable from Big.Contact..Marital.Status
df$Big.Contact..Marital.Status <- factor(
  case_when(
    df$Big.Contact..Marital.Status == "Single" ~ "Single",
    !is.na(df$Big.Contact..Marital.Status) ~ "Not Single",
    TRUE ~ NA_character_
  ),
  levels = c("Single", "Not Single")
)
df$Stage <- ifelse(df$Stage == "Closed", 1, 0)
```

## Understand and analyze the response variable distributions: Match
Length:

```{r}
summary(df$Match.Length)
# Histogram to visualize distribution
hist(df$Match.Length, 
     main="Distribution of Match Length", 
     xlab="Match Length (months)", 
     col="lightblue", 
     breaks=20)

# Density plot
plot(density(df$Match.Length), 
     main="Density Plot of Match Length", 
     xlab="Match Length (months)")

# Boxplot
boxplot(df$Match.Length, 
        main="Boxplot of Match Length", 
        ylab="Match Length (months)")

stats <- summary(df$Match.Length)
mean_value <- mean(df$Match.Length)
q1 <- stats["1st Qu."]
median_value <- stats["Median"]
q3 <- stats["3rd Qu."]

text(x = 1.3, y = mean_value, labels = paste("Mean:", round(mean_value, 2)), col = "black")
text(x = 1.3, y = q1, labels = paste("1Q:", round(q1, 2)), col = "black")
text(x = 1.3, y = median_value, labels = paste("Median:", round(median_value, 2)), col = "black")
text(x = 1.3, y = q3, labels = paste("3Q:", round(q3, 2)), col = "black")

# Check for normality
qqnorm(df$Match.Length)
qqline(df$Match.Length) # Looks exponentially distributed
```

```{r}
table(df$Closure_Reason_Category)
```

## How do the response variable distributions vary across Program Type?

```{r}
table(df$Program.Type)
df$Program.Type[df$Program.Type == "Site Based Plus"] <- NA # too little records to consider
df$Program.Type <- droplevels(df$Program.Type)
df_filtered <- df %>% filter(!is.na(Program.Type))
# Summary statistics by Program Type
summary_stats <- df %>%
  group_by(Program.Type) %>%
  summarise(
    Mean = mean(Match.Length, na.rm = TRUE),
    Median = median(Match.Length, na.rm = TRUE),
    SD = sd(Match.Length, na.rm = TRUE),
    Q1 = quantile(Match.Length, 0.25, na.rm = TRUE),
    Q3 = quantile(Match.Length, 0.75, na.rm = TRUE)
  )
print(summary_stats)

# Boxplot to visualize distribution of Match.Length by Program.Type
ggplot(df_filtered, aes(x = Program.Type, y = Match.Length, fill = Program.Type)) +
  geom_boxplot(na.rm = TRUE) +
  labs(
    title = "Distribution of Match Length by Program Type",
    x = "Program Type",
    y = "Match Length (months)"
  ) +
  theme_minimal()

# Density plot to compare distributions
ggplot(df_filtered, aes(x = Match.Length, fill = Program.Type)) +
  geom_density(alpha = 0.5, na.rm = TRUE) +
  labs(
    title = "Density Plot of Match Length by Program Type",
    x = "Match Length (months)",
    y = "Density"
  ) +
  theme_minimal()

# Violin plot for a more detailed view
ggplot(df_filtered, aes(x = Program.Type, y = Match.Length, fill = Program.Type)) +
  geom_violin(na.rm = TRUE) +
  labs(
    title = "Violin Plot of Match Length by Program Type",
    x = "Program Type",
    y = "Match Length (months)"
  ) +
  theme_minimal()
```

```{r}
# ANOVA test
anova_result <- aov(Match.Length ~ Program.Type, data = df)
summary(anova_result)

# Chi-Square Test of Independence for Closure Reason
chi_square_result <- chisq.test(table(df$Closure_Reason_Category, df$Program.Type))
print(chi_square_result)

# Kruskal-Wallis test (non-parametric alternative)
kruskal.test(Match.Length ~ Program.Type, data = df)
```

Program.Type a significant predictor of Match Length and Closure Reason

## Response distributions over time

```{r}
ggplot(df, aes(x = Match.Activation.Date, y = Match.Length)) +
  geom_point(alpha = 0.5, color = "blue") +  # Scatter plot with transparency
  geom_smooth(method = "loess", color = "red", se = TRUE) +  # Trend line
  labs(title = "Match Length Over Activation Date",
       x = "Activation Date",
       y = "Match Length") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Clearly older matches have an advantage - look into survival analysis
and Cox potential hazards model.

```{r}
ggplot(df, aes(x = Match.Activation.Date, fill = Closure_Reason_Category)) +
  geom_histogram(position = "stack", bins = 30) +
  labs(
    title = "Closure Reason Categories by Match Activation Date",
    x = "Match Activation Date",
    y = "Count",
    fill = "Closure Reason"
  ) +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

df_monthly <- df %>%
  mutate(Month = floor_date(Match.Activation.Date, "month")) %>%
  group_by(Month, Closure_Reason_Category) %>%
  summarise(Count = n(), .groups = "drop") %>%
  group_by(Month) %>%
  mutate(Proportion = Count / sum(Count))

ggplot(df_monthly, aes(x = Month, y = Proportion, fill = Closure_Reason_Category)) +
  geom_area() +
  labs(
    title = "Monthly Proportion of Closure Reasons",
    x = "Match Activation Date (Monthly)",
    y = "Proportion",
    fill = "Closure Reason"
  ) +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

Seems to stay relatively consistent - with a rise in 'NA' values in the
bottom due to less match closures.

```{r}
table(df$Stage)
```

2486 match closures, 789 active matches

```{r}
str(df)
```

## What influence do the various Big and/or Little demographic variables have on the response variable distributions?

```{r}
# Analysis of Big Age vs Match Length
ggplot(df, aes(x = Big.Age, y = Match.Length, color = factor(Stage, labels = c("Active", "Closed")))) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess", color = "blue") +
  labs(
    title = "Match Length by Big's Age",
    x = "Big's Age",
    y = "Match Length (months)",
    color = "Stage"  # Legend title
  ) +
  theme_minimal()

# Age group analysis
df$Age_Group <- cut(df$Big.Age, 
                    breaks = c(0, 25, 35, 45, 55, 65, 100),
                    labels = c("18-25", "26-35", "36-45", "46-55", "56-65", "65+"))

age_summary <- df %>%
  group_by(Age_Group) %>%
  summarise(
    Mean_Length = mean(Match.Length, na.rm = TRUE),
    Median_Length = median(Match.Length, na.rm = TRUE),
    Count = n()
  )
age_summary
```

```{r}
# Big Gender analysis
gender_summary <- df %>%
  group_by(Big.Gender) %>%
  summarise(
    Mean_Length = mean(Match.Length, na.rm = TRUE),
    Median_Length = median(Match.Length, na.rm = TRUE),
    Count = n()
  )
gender_summary
# Box plot of match length by Big's gender
ggplot(df, aes(x = Big.Gender, y = Match.Length, fill = Big.Gender)) +
  geom_boxplot() +
  labs(title = "Match Length by Big's Gender",
       x = "Gender",
       y = "Match Length (months)") +
  theme_minimal()

# Statistical test for gender difference
t.test(Match.Length ~ Big.Gender, data = df)
```

```{r}
# Ethnicity match analysis
ethnicity_summary <- df %>%
  group_by(Ethnicity_Match) %>%
  summarise(
    Mean_Length = mean(Match.Length, na.rm = TRUE),
    Median_Length = median(Match.Length, na.rm = TRUE),
    Count = n()
  )
ethnicity_summary
# Box plot for ethnicity match
ggplot(df %>% filter(!is.na(Ethnicity_Match)), 
       aes(x = factor(Ethnicity_Match), y = Match.Length, fill = factor(Ethnicity_Match))) +
  geom_boxplot() +
  labs(title = "Match Length by Ethnicity Match",
       x = "Ethnicity Match",
       y = "Match Length (months)") +
  scale_x_discrete(labels = c("FALSE" = "No Match", "TRUE" = "Match")) +
  scale_fill_discrete(name = "Ethnicity Match", labels = c("No Match", "Match")) +
  theme_minimal()
```

```{r}
# Interest and proximity analysis
interest_summary <- df %>%
  group_by(has_interests) %>%
  summarise(
    Mean_Length = mean(Match.Length, na.rm = TRUE),
    Median_Length = median(Match.Length, na.rm = TRUE),
    Count = n()
  )
interest_summary
proximity_summary <- df %>%
  group_by(has_proximity) %>%
  summarise(
    Mean_Length = mean(Match.Length, na.rm = TRUE),
    Median_Length = median(Match.Length, na.rm = TRUE),
    Count = n()
  )
proximity_summary
```

```{r}
# County analysis
county_summary <- df %>%
  group_by(County_Factor) %>%
  summarise(
    Mean_Length = mean(Match.Length, na.rm = TRUE),
    Median_Length = median(Match.Length, na.rm = TRUE),
    Count = n()
  ) %>%
  arrange(desc(Mean_Length))
county_summary
# Boxplot for counties
ggplot(df, aes(x = reorder(County_Factor, Match.Length, FUN = median), 
               y = Match.Length, fill = County_Factor)) +
  geom_boxplot() +
  labs(title = "Match Length by County",
       x = "County",
       y = "Match Length (months)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")
```

```{r}
# Occupation analysis
occupation_summary <- df %>%
  group_by(Occupation_Category) %>%
  summarise(
    Mean_Length = mean(Match.Length, na.rm = TRUE),
    Median_Length = median(Match.Length, na.rm = TRUE),
    Count = n()
  ) %>%
  arrange(desc(Mean_Length))
occupation_summary

ggplot(df, aes(x = reorder(Occupation_Category, Match.Length, FUN = median), 
               y = Match.Length, fill = Occupation_Category)) +
  geom_boxplot() +
  labs(title = "Match Length by Big's Occupation",
       x = "Occupation Category",
       y = "Match Length (months)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")

```
Students on average have shorter matches.


```{r}
df_marital <- df %>% filter(!is.na(Big.Contact..Marital.Status))
marital_summary <- df_marital %>%
  group_by(Big.Contact..Marital.Status) %>%
  summarise(
    Mean_Length = mean(Match.Length, na.rm = TRUE),
    Median_Length = median(Match.Length, na.rm = TRUE),
    Count = n()
  )
marital_summary
```


## Analysis of closure reasons by demographics
```{r}
age_closure <- df %>%
  filter(!is.na(Age_Group), !is.na(Closure_Reason_Category)) %>%
  group_by(Age_Group, Closure_Reason_Category) %>%
  summarise(Count = n(), .groups = "drop") %>%
  group_by(Age_Group) %>%
  mutate(Percentage = Count / sum(Count) * 100)

gender_closure <- df %>%
  filter(!is.na(Big.Gender), !is.na(Closure_Reason_Category)) %>%
  group_by(Big.Gender, Closure_Reason_Category) %>%
  summarise(Count = n(), .groups = "drop") %>%
  group_by(Big.Gender) %>%
  mutate(Percentage = Count / sum(Count) * 100)

# Visualize closure reasons by gender
ggplot(gender_closure, aes(x = Big.Gender, y = Percentage, fill = Closure_Reason_Category)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(title = "Closure Reasons by Big's Gender",
       x = "Gender",
       y = "Percentage",
       fill = "Closure Reason") +
  theme_minimal() +
  theme(legend.position = "right")
```

```{r}
# Statistical tests for demographic effects on match length
# ANOVA for County effect
county_anova <- aov(Match.Length ~ County_Factor, data = df)
summary(county_anova)

# ANOVA for Occupation effect
occupation_anova <- aov(Match.Length ~ Occupation_Category, data = df)
summary(occupation_anova)

# ANOVA for Age Group effect
age_anova <- aov(Match.Length ~ Age_Group, data = df)
summary(age_anova)

# Simple regression model
model <- lm(Match.Length ~ Big.Age + Big.Gender + Program.Type + Ethnicity_Match + 
             has_interests + has_proximity + County_Factor + Occupation_Category, 
           data = df %>% filter(!is.na(Ethnicity_Match)))
summary(model)
```
Very poor predictive performance.


## Looking at interests
```{r}
summary_indicators <- df_with_indicators %>%
  summarise(across(c(has_interests, personality_compatibility, has_proximity, 
                    has_commitment, has_experience, has_preference,
                    has_challenges, has_goals), 
                   ~sum(as.integer(as.character(.)) == 1, na.rm = TRUE)))
summary_indicators

# Calculate correlation with match length
indicator_correlations <- df_with_indicators %>%
  select(Match.Length, has_interests, personality_compatibility, has_proximity, 
         has_commitment, has_experience, has_preference,
         has_challenges, has_goals) %>%
  mutate(across(has_interests:has_goals, ~as.numeric(as.character(.)))) %>%
  cor(use = "pairwise.complete.obs")

print(indicator_correlations["Match.Length", ])

# Visualize the distribution of match length by each indicator
library(gridExtra)
```

```{r, fig.width=8, fig.height=12}
# Function to create box plots
create_boxplot <- function(df, var) {
  ggplot(df, aes_string(x = var, y = "Match.Length", fill = var)) +
    geom_boxplot() +
    labs(title = paste("Match Length by", var),
         x = var,
         y = "Match Length (months)") +
    theme_minimal() +
    theme(legend.position = "none") +
    scale_x_discrete(labels = c("0" = "No", "1" = "Yes"))
}

# Create a list of plots
plot_list <- lapply(c("has_interests", "personality_compatibility", "has_proximity", 
                      "has_commitment", "has_experience", "has_preference",
                      "has_challenges", "has_goals"), 
                    function(var) create_boxplot(df_with_indicators, var))

# Arrange plots in a grid
grid.arrange(grobs = plot_list, ncol = 2)
```

```{r, fig.width=4, fig.height=4}
# Analyze the impact of different interest combinations
df_with_indicators$interest_count <- rowSums(sapply(df_with_indicators[, c("has_interests", 
                                                                         "personality_compatibility", 
                                                                         "has_proximity", 
                                                                         "has_commitment", 
                                                                         "has_experience", 
                                                                         "has_preference",
                                                                         "has_challenges", 
                                                                         "has_goals")], 
                                                  function(x) as.integer(as.character(x))))

# Analyze relationship between number of interest categories and match length
interest_count_summary <- df_with_indicators %>%
  group_by(interest_count) %>%
  summarise(
    Mean_Length = mean(Match.Length, na.rm = TRUE),
    Median_Length = median(Match.Length, na.rm = TRUE),
    Count = n()
  )

print(interest_count_summary)

# Visualize relationship between interest count and match length
ggplot(df_with_indicators, aes(x = factor(interest_count), y = Match.Length)) +
  geom_boxplot(fill = "lightblue") +
  labs(title = "Match Length by Number of Interest Categories Mentioned",
       x = "Number of Interest Categories",
       y = "Match Length (months)") +
  theme_minimal()

# Test statistical significance
interest_model <- lm(Match.Length ~ has_interests + personality_compatibility + has_proximity + 
                       has_commitment + has_experience + has_preference +
                       has_challenges + has_goals, data = df_with_indicators)
summary(interest_model)
```

## Successful match?
```{r}
# A match is successful if:
# 1. It's active 
# 2. It has a long duration 
# 3. Closure reason is "Success"

df$successful_match <- FALSE

# Active matches (no closure date)
df$successful_match[df$Stage == 0] <- TRUE
sum(df$successful_match[df$Stage == 0])

# Matches with "Success" as closure reason
df$successful_match[df$Closure_Reason_Category == "Success"] <- TRUE

# Long duration matches (define long as above the 75th percentile)
long_duration_threshold <- quantile(df$Match.Length, 0.75, na.rm = TRUE)
df$successful_match[df$Match.Length > long_duration_threshold & 
                      !is.na(df$Match.Length)] <- TRUE

# Success rate
success_rate <- mean(df$successful_match, na.rm = TRUE)
print(paste("Overall success rate:", round(success_rate * 100, 1), "%"))
```


analyzing factors associated with successful matches
```{r}
# 1. Mentor age group
age_group_success <- aggregate(successful_match ~ Age_Group, data = df, FUN = mean)
age_group_success$count <- aggregate(successful_match ~ Age_Group, data = df, FUN = length)$successful_match
age_group_success <- age_group_success[order(-age_group_success$successful_match),]
print("Success rate by mentor age group:")
print(age_group_success)

# 2. Gender
gender_success <- aggregate(successful_match ~ Big.Gender, data = df, FUN = mean)
gender_success$count <- aggregate(successful_match ~ Big.Gender, data = df, FUN = length)$successful_match
print("Success rate by mentor gender:")
print(gender_success)

# 3. Program type
program_type_success <- aggregate(successful_match ~ Program.Type, data = df, FUN = mean)
program_type_success$count <- aggregate(successful_match ~ Program.Type, data = df, FUN = length)$successful_match
print("Success rate by program type:")
print(program_type_success)

# 4. Ethnicity match
ethnicity_match_success <- aggregate(successful_match ~ Ethnicity_Match, data = df, FUN = mean)
ethnicity_match_success$count <- aggregate(successful_match ~ Ethnicity_Match, data = df, FUN = length)$successful_match
print("Success rate by ethnicity match:")
print(ethnicity_match_success)

# 5. Occupation category
occupation_success <- aggregate(successful_match ~ Occupation_Category, data = df, FUN = mean)
occupation_success$count <- aggregate(successful_match ~ Occupation_Category, data = df, FUN = length)$successful_match
occupation_success <- occupation_success[order(-occupation_success$successful_match),]
print("Success rate by occupation category:")
print(occupation_success)

# 6. County
county_success <- aggregate(successful_match ~ County_Factor, data = df, FUN = mean)
county_success$count <- aggregate(successful_match ~ County_Factor, data = df, FUN = length)$successful_match
county_success <- county_success[order(-county_success$successful_match),]
print("Success rate by county:")
print(county_success)

# 7. Compatibility factors
# Create a function to check success rate for binary factors
check_binary_factor <- function(factor_name) {
  formula <- as.formula(paste("successful_match ~", factor_name))
  success_rate <- aggregate(formula, data = df, FUN = mean)
  success_rate$count <- aggregate(formula, data = df, FUN = length)$successful_match
  print(paste("Success rate by", factor_name, ":"))
  print(success_rate)
}

compatibility_factors <- c("has_interests", "personality_compatibility", "has_proximity", 
                          "has_commitment", "has_experience", "has_preference",
                          "has_challenges", "has_goals")

for (factor in compatibility_factors) {
  check_binary_factor(factor)
}

# 8. Logistic regression to identify key predictors of success
# First, convert factors to proper format where needed
df$successful_match_numeric <- as.numeric(df$successful_match)

model <- glm(successful_match_numeric ~ Big.Age + Big.Gender + Program.Type + 
              Ethnicity_Match + County_Factor + Occupation_Category + Age_Group +
              has_interests + personality_compatibility + has_proximity + 
              has_commitment + has_experience + has_preference + has_challenges + has_goals,
            family = binomial(link = "logit"), data = df)

# Print summary of model
summary_model <- summary(model)
print("Logistic regression results (key predictors of successful matches):")
print(summary_model)

# Simplified output of key predictors (those with significant p-values)
significant_predictors <- summary_model$coefficients[summary_model$coefficients[,4] < 0.05,]
print("Significant predictors of match success:")
print(significant_predictors)

# 9. Key insights and recommendations
print("Key insights and recommendations for Big Brothers Big Sisters Twin Cities:")
print("1. Most important factors for successful matches:")
```


